\Title

### 2012-12-21 synmut.tex
small changes here and there

\begin{abstract}
\noindent
Intrapatient HIV evolution is goverened by selection on the protein level in the
arms race with the immune system (killer T-cells and antibodies). Synonymous
mutations do not have an immunity-related phenotype and are often assumed to be
neutral. In this paper, we show that synonymous changes in epitope-rich regions
are often deleterious but still reach frequencies of order one.  We analyze time
series of viral sequences from the V1-C5 part of {\it env} within individual
hosts and observe that synonymous derived alleles rarely fix in the
viral population. Simulations suggest that such synonymous mutations
have a (Malthisuan) selection coefficient of the order of $-0.001$, and that
they are brought up to high frequency by linkage to neighbouring beneficial
nonsynonymous alleles (genetic draft). As far as the biological causes are
concerned, we detect a negative correlation between fixation of an allele and
its involvement in evolutionarily conserved RNA stem-loop structures.
This phenonenon is not observed in other parts of the HIV genome, in which
selective sweeps are less dense and the genetic architecture less constrained.
\end{abstract}

\maketitle

\section{Introduction}
HIV evolves rapidly within a single host during the course of the infection.
This evolution is driven by strong selection imposed by the host immune system
via killer T cells (CTLs) and neutralizing antibodies
(AB)~\citep{pantaleo_immunopathogenesis_1996} and facilitated by the high
mutation rate of HIV~\citep{mansky_lower_1995}. When the host develops a CTL or
AB response against a particular HIV epitope, mutations in the viral genome that reduce or prevent
recognition of the epitope frequently emerge. Escape mutations in epitopes
targeted by CTLs typically evolve during early infection and spread rapidly through
the population~\citep{mcmichael_immune_2009}. During chronic infection, the most
rapidly evolving part of the HIV genome are the so called variable loops of the
envelope protein gp120, which need to avoid recognition by neutralizing ABs.
Mutations in \env~, the gene encoding for gp120, spread through the population
within a few months (see \figurename~\ref{fig:aft}, solid lines).
The (Malthusian) effect size of these beneficial mutations is of the
order of $s_a \sim 0.01$~\citep{neher_recombination_2010}.

These escape mutations are strongly selected for their effect on the amino acid
sequence of the viral proteins. Conversely, synonymous mutations are commonly
used as approximate neutral markers in studies of viral evolution. Neutral
markers are very useful in practice, because they can be used to make inferences
about the stochastic forces driving evolution~\citep{yang_statistical_2000}.
The viral genome, however, needs to satisfy further constraints in addition to
immune escape, such as efficient processing and translation, nuclear export, and
packaging into the viral capsid: all these processes operate at the RNA level
and are sensitive to synonymous changes. A
few functionally important RNA elements are well characterized. For example, a certain RNA
sequence, called \rev{} response element (RRE), is used by HIV to enhance
nuclear export of some of its transcripts~\citep{fernandes_hiv-1_2012}. Another
well studied case is the interaction between viral reverse transcriptase, viral
ssRNA, and the host tRNA$^\text{Lys3}$: the latter is required for priming
reverse transcription (RT) and bound by a specifical pseudoknotted RNA structure
in the viral 5' untranslated region~\citep{barat_interaction_1991,
paillart_vitro_2002}. Furthermore, recent studies have shown that genetically
engineered HIV strains with skewed codon usage bias (CUB) patterns towards more
or less abundant tRNAs replicate better or worse,
respectively~\citep{ngumbela_quantitative_2008, li_codon-usage-based_2012}.
Purifying selection beyond the protein sequence is therefore expected, while it
seems reasonable that the bulk of positive selection through the immune system
should be restricted to amino acid sequences.

INFLUENZA PSEUDO VACCINE.

@@

Here, we characterize the dynamics of synonymous mutations in \env{} and show
that a substantial fraction of these mutations are deleterious. We further show
that, although such synonymous mutations cannot be used as neutral markers, the
degree to which they hitchhike with nearby nonsynonymous mutations is very
informative. Their ability to hitchhike for extended times, which is a core
requirement for our analysis, is rooted in the small recombination rate of
HIV~\citep{neher_recombination_2010, batorsky_estimate_2011}. Extending the
analysis of fixation probabilities to the nonsynonymous mutations, we show that
time dependent selection or strong competition of escape mutations inside the
same epitope are necessary to explain the observed patterns of fixation and
loss.

\section{Results}
The central quantity we investigate is the probability of fixation
of a mutation, conditional on its population frequency.
A neutral mutation segregating at frequency $\nu$ has a probability $\nu$ to
spread through the population and fix; in the rest of the cases, i.e. with probability
$1-\nu$, it goes extinct. This is a simple consequence of the fact that (a) exactly
one of the $N$ individuals in the current population will be the common ancestor
of the entire future population at a
particular locus and (b) this ancestor has a probability $\nu$ of carrying the
mutation, see illustration in \FIG{fixp}. Deleterious or beneficial
mutations fix less or more often than neutral ones, respectively. Time series
sequence data enable a direct observation of both the current frequency $\nu$ of
any particular mutation and its future fate (fixation or extinction).
They therefore represent a simple way to investigate average properties
of different classes of mutations.

@@

\FIG{aft} shows time series data of the frequencies of all mutations observed
\env~, C2-V5, in patient p8~\citep{shankarappa_consistent_1999,
liu_selection_2006}. Despite many synonymous mutations reaching high frequency,
very few fix. This observation is quantified in panels \ref{fig:fixp1} and
\ref{fig:fixp2}, which stratify the data of 7-10 patients according to the
frequency at which different mutations are observed (see methods). Considering
all mutations in a frequency interval around $\nu_0$ at some time $t_i$, we
calculate the fraction that is found at frequency 1, at frequency 0, or at
intermediate frequency at later times $t_f$. Plotting these fixed, lost, and
polymorphic fraction against the time interval $t_f-t_i$, we see that most
synonymous mutations segregate for roughly one year and are lost much more
frequently than expected. The long-time probability of fixation versus
extinction is shown as a function of the initial frequency $\nu_0$ in
panel~\ref{fig:fixp2}. In contrast to synonymous mutations, the nonsynonymous
seem to follow more a less the neutral expectation -- a point to which we will
come back below. 
\begin{figure}
\begin{center}
\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_syn_nonsynp8}
\caption{Synonymous mutations rarely fix in \env, C2-V5: mutation frequency
 trajectories observed in patient 8~\cite{shankarappa_consistent_1999};
 Nonsynonymous and synonymous mutations are shown as solid and dashed lines,
 respectively. Colors indicate the position of the site along the C2-V5 region
 (red to blue) MAYBE MAKE FIGURE WITH SYNONYMOUS AND NONSYN
 SEPARATELY. While nonsynonymous mutations frequently fix, very few synonymous
mutations do even though they are frequently observed at intermediate
frequencies.}
\label{fig:aft}
\end{center}
\end{figure}

\citet{bunnik_autologous_2008} present a longitudinal dataset on the entire
\env~gene of 3 patients at $\sim 5$ time points with approximately 5-20
sequences each (see methods). Repeating the above analysis separately on the C2-V5 region
studied above and the remainder of \env~ reveal strikingly different behavior
inside and outside the hypervariable region. Within C2-V5, this data fully
confirms the observations made in the data set by
\citet{shankarappa_consistent_1999}. In the remainder of \env, however, observed
synonymous mutations behave as if they were neutral; see \FIG{fixp}.

%ARE OBSERVED SYNONYMOUS MUTATIONS OUTSIDE C2-V5 NEUTRAL? (?? SOME!)
%DOES LOSS/FIX CORRELATE WITH CONSERVATION? YES.
%MAYBE WE COULD HAVE ONE -- COMPLETELY CIRCULAR -- FIGURE SHOWING LOSS/FIX VS CONSERVATION: YES.
%CAN WE LOOK AT THE AVERAGE LEVEL OF CONSERVATION STRATIFIED BY MAX FREQ? TRICKY: the maximal freq is achieved by hitchhiking...

@@

\subsection{Synonymous mutations in C2-V5 tend to disrupt conserved RNA stems}
One possible {\it a priori} explanation for lack of fixation of synonymous
mutations in C2-V5 are  secondary structures in the viral RNA. If any RNA
secondary structures are relevant for HIV replication, mutations in nucleotides
involved in those base pairs are expected to be deleterious and to revert
preferentially. Many functionally important secondary structure elements have
been characterized, including  the RRE~\citep{fernandes_hiv-1_2012} the
5' UTR pseudoknot interacting with thehost tRNA$^\text{Lys3}$~\citep{barat_interaction_1991,
paillart_vitro_2002}. It has been suggested early on that parts of the viral
genome that has the potential to form stems is better conserved that the
remainder~\citep{forsdyke_reciprocal_1995}.

Recently, the propensity of nucleotides of the HIV genome to form base pairs has
been measured using the SHAPE assay (a biochemical reaction preferentially
altering unpaired bases)~\citep{watts_architecture_2009}. The SHAPE assay has
shown that the variable regions V1 to V5 tend to be unpaired, while the
conserved regions between those variable regions form stems. We partition all
synonymous alleles observed at intermediate frequencies above 10-15\% depending
on their final destiny (fixation or extinction). Subsequently, we align our
sequences to the reference NL4-3 strain used in
ref.~\citep{watts_architecture_2009} and assign them SHAPE reactivities. As
shown in \FIG{SHAPEA} in a cumulative histogram, the reactivity of fixed alleles
are systematically larger than of alleles that are doomed to extinction
(Kolmogorov-Smirnov test, $P\approx 2~\text{\textperthousand}$). In other
words, alleles that are likely to be breaking RNA helices are also more likely
to revert and finally be lost from the population.  As a complementary analysis,
we split the synonymous mutations in the C2-V5 region further into conserved and
variable regions and found that the biggest depression in fixation probability
is observed in the conserved stems, while the variable loops show little
deviation from the neutral signature, see \FIG{SHAPEB}.

@@

\subsection{Deleterious mutations are brought to high frequency by hitch-hiking}
While the observation that some fraction of synonymous mutations is deleterious
is not unexpected, it seems odd that we observe them at high population
frequency -- at least in some regions of the genome. The region of \env~ in which
we observe deleterious mutations at high frequency is special in that it
undergoes frequent adaptive changes to evade recognition by neutralizing
antibodies~\cite{williamson_adaptation_2003}. Due to the limited amount of
recombination in HIV~\cite{neher_recombination_2010,batorsky_estimate_2011},
deleterious mutations that are linked to adaptive variants can reach high
frequency~\citep{smith_hitch-hiking_1974}.

@@

To get a better idea of the range of parameters that are compatible with the
observations and our interpretation, we  perform computer simulations of
evolving viral populations under selection and rare recombination. For this
purpose, we use the recently published package FFPopSim, which includes a module
dedicated to intrapatient HIV evolution~\citep{zanini_ffpopsim:_2012}. We
analyze many combinations of parameters such as population size, recombination
rate, selection coefficient and density of escape mutations, deleterious effect
of synonymous mutation.

The main result of the simulations is that genetic draft can indeed bring weakly
deleterious mutations to high frequencies and result in a dependence of the
fixation probability on initial frequency that is compatible with observations.
First of all, since neutral mutations
are much more likely to rise to high frequency than deleterious ones, the
majority of the synonymous mutations needs to be slightly deleterious observe a
significant reduction of $P_\text{fix}$.
In order to further quantify the reduction in fixation probability, we look at the
difference between the neutral curve ($P(\nu) = \nu$) and the measured fixation
probability and calculate its area (see inset of \FIG{simfixpvar}). The minimal and maximal values for this
area are zero (neutral-like curve) and 0.5 (no fixation at all),
respectively. The convex curve seen in the HIV data corresponds to an area of
approximately 0.17. Various simulation curves are shown in \FIG{simfixpvar}, and
the area of the data curve is shown in the inset.
In \FIGS{simheat1}{simheat2}, we explore the parameter space: the combinations that yield areas close to the
experimental result are roughly indicated by ellipses. The two crucial parameters
that control the fixation probability are the following: (a) the deleterious
effects of hitchhikers compared to the beneficial effects of escape mutants, and
(b) the density of escape mutations. Intuitively, a higher density of escape
mutations (i.e., epitopes) enables a larger degree of genetic draft, because
escape mutations start to combine and their effects add up. In \FIG{simheat1},
we show that this is indeed the case in simulations.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.9\linewidth]{fixation_loss_shortgenome_distance_ada_frac_del_eff_coi_various.pdf}
\label{fig:simfixpvar}}\\
\subfloat{\includegraphics[width=0.9\linewidth]{fixation_loss_shortgenome_area_ada_frac_del_eff_coi_0_01_nescepi_6_heat.pdf}
\label{fig:simheat1}}\\
\subfloat{\includegraphics[width=0.9\linewidth]{fixation_loss_shortgenome_area_ada_frac_del_eff_coi_0_01_nescepi_6_nonsyn_heat.pdf}
\label{fig:simheat2}}
\caption{The depression in $P_\text{fix}$ depends on the deleterious effect size
 of the synonymous alleles (panel A). Simulations on the escape competition
 scenario show that the density of selective sweeps and the size of the
 deleterious effects of synonymous mutations are the main driving forces of the
 phenomenon. A convex fixation probability is recovered, as seen in the data,
 along the diagonal (panel B): more dense sweeps can support more deleterious
 linked mutations. The density of sweeps is limited, however, by the
 nonsynonymous fixation probability, which is quite close to neutrality (panel
 C). Moreover, strong competition between escape mutants is required, so that
 several escape mutants are ``found'' by HIV within a few months of antibody
production.}
\label{fig:simheat}
\end{center}
\end{figure}

However, if hitch-hiking is driven by nonsynonymous mutations that are
unconditionally beneficial, we should find that nonsynonymous mutations almost
always fix once they reach high frequencies -- in contrast with \FIG{fixp} that
shows that nonsynonymous mutations fix as if they were neutral. We know,
however, that nonsynonymous variation in the variable regions is driven by
positive selection. Inspecting the trajectories of nonsynonymous mutations
suggest the rapid rise and fall of many alleles.  We test two possible such
mechanisms that are biologically plausible and could explain the transient rise
of nonsynonymous mutations: time-dependent selection and within-epitope
competition. If the immune system recognizes the escape mutant before its
fixation, the mutant might cease to be beneficial and disappear despite its
quick initial rise in frequency.  In support of this idea,
\citet{richman_rapid_2003, bunnik_autologous_2008} report antibody responses to
escape mutants. These respones are delayed by a few months, roughly matching the
average sweep time of an escape mutant. Alternatively, several different escape
mutations in the same epitope can arise almost simultaneously and start to
spread. Their fitness benefits are not additive, because each of them is
essentially sufficient to escape. As a consequence, several escape mutations rise to
high frequency, while the escape with the smallest cost in terms of replication,
packaging, etc. is most likely to
eventually fix. In simulations, this kind of epistatic interactions within
epitopes reduces the fixation probability. The emergence of
multiple sweeping nonsynonymous mutations in real HIV infections has been shown
previously~\citep{moore_limited_2009, bar_early_2012}.
See the supplementary material for examples of successful simulations in both scenarios.

@@

The observed hitchhiking highlights the importance of linkage due to infrequent
recombination for the evolution of HIV
\citep{neher_recombination_2010,batorsky_estimate_2011,
josefsson_majority_2011}. The recombination rate has been estimated to be on the
order of $\rho = 10^{-5}$ per base and day. It takes roughly $t_{sw} = s_a^{-1}
\log \nu_0$ generations for an adaptive mutation with growth rate $s_a$ to rise
from an initially low frequency $\nu_0\sim \mu$ to frequency one. This implies
that a region of length $l = (\rho t_{sw})^{-1} = s_a / \rho \log \nu_0$ remains
linked to the adaptive mutation. With $s_a=0.01$, $l\approx 100$ bases which is
consistent with strong linkage between the variable loops and the stems in
between. Furthermore, we do not expect hitchhiking to extend far beyond
the variable regions consistent with the lack of signal out side of C5-V5. In
case of much stronger selection -- such as observed during early CTL escape or
drug resistance evolution -- the linked  region is of course much larger.

The functional significance of the insulating RNA structure stems between the
hyper variable loops has been proposed
previously~\citep{watts_architecture_2009, sanjuan_interplay_2011}.
\citet{sanjuan_interplay_2011} have shown that insulating stems are relevant for
viral fitness {\it in vivo}. Our analysis is limited by the availability of
longitudinal data which requires a focus on the the variable regions of \env~.
Conserved RNA structures most likely exist in different parts
of the HIV genome (several are known). In absence of repeated adaptive substitutions in the vicinity
that cause hitchhiking, the deleterious synonymous mutations remain at low
frequencies and can only be observed by deep sequencing methods.

>>>>

\begin{abstract}
\noindent

Intrapatient HIV evolution is goverened by selection on the protein level in
the arms race with the immune system (killer T-cells and antibodies).
Synonymous mutations do not have an immunity-related phenotype and are often
assumed to be neutral. In this paper, we show that synonymous changes in
epitope-rich regions are often deleterious but reach frequencies of order one
via genetic hitchhiking.  We analyze time series of viral sequences from the
V1-C5 part of {\it env}, the envelope gene, within individual hosts and observe
that synonymous derived alleles fix in the viral population much less
frequently than expected from neutral models. Instead, synonymous changes tend
to revert on a time scale of two years, suggesting a selection coefficient of the
order of $-1~$\textperthousand. Extensive computer simulations support these
findings quantitatively. We explore possible biological causes and detect a
negative correlation between fixation of an allele and its involvement in
secondary RNA stem-loop structures, indicating that such structures are
functionally relevant for HIV replication {\it in vivo}. This phenonenon is not
observed in other parts of the HIV genome, in which selective sweeps are less
dense and the genetic architecture less constrained.

\end{abstract}
\maketitle

\section{Introduction}

HIV evolves rapidly within a single host during the course of the infection.
This evolution is driven by strong selection imposed by the host immune system
via killer T cells (CTLs) and neutralizing antibodies
(ABs)~\citep{pantaleo_immunopathogenesis_1996} and facilitated by the high
mutation rate of HIV~\citep{mansky_lower_1995}. When the host develops a CTL or
AB response against a particular HIV epitope, mutations in the viral genome that
reduce or prevent recognition of the epitope frequently emerge. Escape mutations
in epitopes targeted by CTLs typically evolve during early infection and spread
rapidly through the population~\citep{mcmichael_immune_2009}. During chronic
infection, the most rapidly evolving part of the HIV genome are the so called
variable loops of the envelope protein gp120, which need to avoid recognition by
neutralizing ABs.  Mutations in \env~, the gene encoding for gp120, spread
through the population within a few months (see \figurename~\ref{fig:aft}, solid
lines).  The (Malthusian) effect size of these beneficial mutations is of the
order of $s_a \sim 0.01$~\citep{neher_recombination_2010}.

These escape mutations are strongly selected for their effect on the amino acid
sequence of the viral proteins. Conversely, synonymous mutations are commonly
used as approximate neutral markers in studies of viral evolution. Neutral
markers are very useful in practice, because they can be used to make inferences
about the stochastic forces driving evolution~\citep{yang_statistical_2000}.
The viral genome, however, needs to satisfy further constraints in addition to
immune escape, such as efficient processing and translation, nuclear export, and
packaging into the viral capsid: all these processes operate at the RNA level
and are sensitive to synonymous changes. A few functionally important RNA
elements are well characterized. For example, a certain RNA sequence in the HIV
genome, called \rev{} response element (RRE), enhances nuclear export of viral
transcripts~\citep{fernandes_hiv-1_2012}. Another well studied case is the
interaction between viral reverse transcriptase, viral ssRNA, and the host
tRNA$^\text{Lys3}$: the latter is required for priming reverse transcription
(RT) and bound by a specifical pseudoknotted RNA structure in the viral 5'
untranslated region~\citep{barat_interaction_1991, paillart_vitro_2002}.
Nucleotide-level fitness effects have been observed beyond RNA structure as
well. Recent studies have shown that genetically engineered HIV strains with
skewed codon usage bias (CUB) patterns towards more or less abundant tRNAs
replicate better or worse, respectively~\citep{ngumbela_quantitative_2008,
li_codon-usage-based_2012}. A similar conclusion has been reached about
influenza, and codon-pessimized influenza strains have been shown to be good
live attenuated vaccines in mice~\citep{mueller_live_2010}. Purifying selection
beyond the protein sequence is therefore expected, while it seems reasonable
that the bulk of positive selection through the immune system be restricted to
amino acid sequences.

@@

In this paper, we characterize the dynamics of synonymous mutations in \env{}
and show that a substantial fraction of these mutations is deleterious. We
further show that, although such synonymous mutations cannot be used as neutral
markers, the degree to which they hitchhike with nearby nonsynonymous mutations
is very informative. Their ability to hitchhike for extended times, which is a
core requirement for our analysis, is rooted in the small recombination rate of
HIV~\citep{neher_recombination_2010, batorsky_estimate_2011}. Extending the
analysis of fixation probabilities to the nonsynonymous mutations, we show that
time dependent selection or strong competition of escape mutations inside the
same epitope are necessary to explain the observed patterns of fixation and
loss.

\section{Results}

The central quantity we investigate is the probability of fixation of a
mutation, conditional on its population frequency.  A neutral mutation
segregating at frequency $\nu$ has a probability $P_\text{fix}(\nu) = \nu$ to
spread through the population and fix; in the rest of the cases, i.e. with
probability $1-\nu$, it goes extinct. This is a simple consequence of the fact
that (a) exactly one of the $N$ individuals in the current population will be
the common ancestor of the entire future population at a particular locus and
(b) this ancestor has a probability $\nu$ of carrying the mutation, see
illustration in \FIG{fixp}.  Deleterious or beneficial mutations fix less or
more often than neutral ones, respectively. Time series sequence data enable a
direct observation of both the current frequency $\nu$ of any particular
mutation and its future fate (fixation or extinction). They therefore represent
a simple way to investigate average properties of different classes of
mutations.

@@

\FIG{aft} shows time series data of the frequencies of all mutations observed
\env~, C2-V5, in patient p8~\citep{shankarappa_consistent_1999}. Despite many
synonymous mutations reaching high frequency (dashed lines), very few fix. This
observation is quantified in panels \FIG{fixp1} and \ref{fig:fixp2}, which
stratify the data of 7 (resp. 10) patients according to the frequency at which
different mutations are observed (see methods). Considering all mutations in a
frequency interval around $\nu_0$ at some time $t_i$, we calculate the fraction
that is found at frequency 1, at frequency 0, or at intermediate frequency at
later times $t_f$. Plotting these fixed, lost, and polymorphic fraction against
the time interval $t_f-t_i$, we see that most synonymous mutations segregate for
roughly two years and are lost much more frequently than expected. The long-time
probability of fixation versus extinction is shown as a function of the initial
frequency $\nu_0$ in panel~\ref{fig:fixp2} (red line). In contrast to synonymous
mutations, the nonsynonymous seem to follow more a less the neutral expectation
(blue line) -- a point to which we will come back below.

\begin{figure}
\begin{center}
\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_syn_nonsynp8}
\caption{Synonymous mutations rarely fix in \env, C2-V5: mutation frequency
 trajectories observed in patient 8~\cite{shankarappa_consistent_1999};
 Nonsynonymous and synonymous mutations are shown as solid and dashed lines,
 respectively. Colors indicate the position of the site along the C2-V5 region
 (red to blue) MAYBE MAKE FIGURE WITH SYNONYMOUS AND NONSYN
 SEPARATELY. While nonsynonymous mutations frequently fix, very few synonymous
mutations do even though they are frequently observed at intermediate
frequencies.}
\label{fig:aft}
\end{center}
\end{figure}

\citet{bunnik_autologous_2008} present a longitudinal dataset on the entire
\env~gene of 3 patients at $\sim 5$ time points with approximately 5-20
sequences each (see methods). Repeating the above analysis separately on the
C2-V5 region studied above and the remainder of \env~ reveal striking
differences (see \FIG{fixp}). Within C2-V5, this data fully confirms the
observations made in the data set by \citet{shankarappa_consistent_1999} (red
line). In the remainder of \env, however, observed synonymous mutations behave
as if they were neutral (orange line).

%ARE OBSERVED SYNONYMOUS MUTATIONS OUTSIDE C2-V5 NEUTRAL? (?? SOME!)
%DOES LOSS/FIX CORRELATE WITH CONSERVATION? YES.
%MAYBE WE COULD HAVE ONE -- COMPLETELY CIRCULAR -- FIGURE SHOWING LOSS/FIX VS CONSERVATION: SUPPLEMENTARY?
%CAN WE LOOK AT THE AVERAGE LEVEL OF CONSERVATION STRATIFIED BY MAX FREQ? TRICKY: the maximal freq is achieved by hitchhiking...

@@

\subsection{Synonymous mutations in C2-V5 tend to disrupt conserved RNA stems}

One possible {\it a priori} explanation for lack of fixation of synonymous
mutations in C2-V5 are  secondary structures in the viral RNA. If any RNA
secondary structures are relevant for HIV replication, mutations in nucleotides
involved in those base pairs are expected to be deleterious and to revert
preferentially. Many functionally important secondary structure elements have
been characterized, including  the RRE~\citep{fernandes_hiv-1_2012} and the 5'
UTR pseudoknot interacting with the host
tRNA$^\text{Lys3}$~\citep{barat_interaction_1991, paillart_vitro_2002}. It has
been suggested early on that parts of the viral genome that has the potential to
form stems is better conserved than the
remainder~\citep{forsdyke_reciprocal_1995}.

Recently, the propensity of nucleotides of the HIV genome to form base pairs has
been measured using the SHAPE assay (a biochemical reaction preferentially
altering unpaired bases)~\citep{watts_architecture_2009}. The SHAPE assay has
shown that the variable regions V1 to V5 tend to be unpaired, while the
conserved regions between those variable regions form stems. We partition all
synonymous alleles observed at intermediate frequencies above 10-15\% depending
on their final destiny (fixation or extinction). Subsequently, we align our
sequences to the reference NL4-3 strain used in
ref.~\citep{watts_architecture_2009} and assign them SHAPE reactivities. As
shown in \FIG{SHAPEA} in a cumulative histogram, the reactivities of fixed
alleles (red line) are systematically larger than of alleles that are doomed to
extinction (blue line) (Kolmogorov-Smirnov test, $P\approx
2~\text{\textperthousand}$). In other words, alleles that are likely to be
breaking RNA helices are also more likely to revert and finally be lost from the
population. As a control, the average over non-observed but potentially
available polymorphisms lies between the two curves (green line), as expected
(because only some of them will be helix breakers). Furthermore, as a
complementary analysis, we split the synonymous mutations in the extended V1-V5
region further into conserved and variable regions and found that the biggest
depression in fixation probability is observed in the conserved stems, while the
variable loops show little deviation from the neutral signature, see
\FIG{SHAPEB}.

@@

\subsection{Deleterious mutations are brought to high frequency by hitch-hiking}

While the observation that some fraction of synonymous mutations is deleterious
is not unexpected, it seems odd that we observe them at high population
frequency -- at least in some regions of the genome. The region of \env~ in
which we observe deleterious mutations at high frequency, however, is special in
that it undergoes frequent adaptive changes to evade recognition by neutralizing
antibodies~\cite{williamson_adaptation_2003}. Due to the limited amount of
recombination in HIV~\cite{neher_recombination_2010,batorsky_estimate_2011},
deleterious mutations that are linked to adaptive variants can reach high
frequency~\citep{smith_hitch-hiking_1974}.

@@

To get a better idea of the range of parameters that are compatible with the
observations and our interpretation, we  perform computer simulations of
evolving viral populations under selection and rare recombination. For this
purpose, we use the recently published package FFPopSim, which includes a module
dedicated to intrapatient HIV evolution~\citep{zanini_ffpopsim:_2012}. We
analyze many combinations of parameters such as population size, recombination
rate, selection coefficient and density of escape mutations, and deleterious effect
of synonymous mutation.

The main result of the simulations is that genetic draft can indeed bring weakly
deleterious mutations to high frequencies and result in a dependence of the
fixation probability on initial frequency that is compatible with observations.
First of all, since neutral mutations are much more likely to rise to high
frequency than deleterious ones, the majority of the synonymous mutations needs
to be slightly deleterious observe a significant reduction of $P_\text{fix}$.
In order to further quantify the reduction in fixation probability, we look at
the difference between the neutral curve ($P_\text{fix}(\nu) = \nu$) and the
measured fixation probability and calculate its area (see inset of
\FIG{simfixpvar}). The minimal and maximal values for this area are zero
(neutral-like curve) and 0.5 (no fixation at all), respectively. The HIV data
correspond to an area under the diagonal of approximately 0.2 for synonymous
changes, and a very small area over the diagonal for nonsynonymous changes.
Various simulation curves are shown in \FIG{simfixpvar}. Then, in
\FIGS{simheat1}{simheat2}, we explore the parameter space: the combinations that
yield areas close to the experimental result are roughly indicated by ellipses.
The two crucial parameters that control the fixation probability are the
following: (a) the deleterious effects of hitchhikers compared to the beneficial
effects of escape mutants, and (b) the density of escape mutations. Intuitively,
a higher density of escape mutations (i.e., epitopes) enables a larger degree of
genetic draft, because escape mutations from different epitopes start to combine
and their effects add up. In \FIG{simheat1}, we show that this is indeed the
case in simulations.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.9\linewidth]{fixation_loss_shortgenome_distance_ada_frac_del_eff_coi_various.pdf}
\label{fig:simfixpvar}}\\
\subfloat{\includegraphics[width=0.9\linewidth]{fixation_loss_shortgenome_area_ada_frac_del_eff_coi_0_01_nescepi_6_heat.pdf}
\label{fig:simheat1}}\\
\subfloat{\includegraphics[width=0.9\linewidth]{fixation_loss_shortgenome_area_ada_frac_del_eff_coi_0_01_nescepi_6_nonsyn_heat.pdf}
\label{fig:simheat2}}
\caption{\comment{USE A GOOD FIG WITH THE CORRECT $s_a$!} The depression in $P_\text{fix}$ depends on the deleterious effect size
 of the synonymous alleles (panel A). Simulations on the escape competition
 scenario show that the density of selective sweeps and the size of the
 deleterious effects of synonymous mutations are the main driving forces of the
 phenomenon. A convex fixation probability is recovered, as seen in the data,
 along the diagonal (panel B): more dense sweeps can support more deleterious
 linked mutations. The density of sweeps is limited, however, by the
 nonsynonymous fixation probability, which is quite close to neutrality (panel
 C). Moreover, strong competition between escape mutants is required, so that
 several escape mutants are ``found'' by HIV within a few months of antibody
production.}
\label{fig:simheat}
\end{center}
\end{figure}

However, if hitchhiking is driven by nonsynonymous mutations that are
unconditionally beneficial, we should find that nonsynonymous mutations almost
always fix once they reach high frequencies -- in contrast with \FIG{fixp} that
shows that nonsynonymous mutations fix as if they were neutral. We know,
however, that nonsynonymous variation in the variable regions is driven by
positive selection. Inspecting the trajectories of nonsynonymous mutations
suggest the rapid rise and fall of many alleles.  We test two possible
mechanisms that are biologically plausible and could explain the transient rise
of nonsynonymous mutations: time-dependent selection and within-epitope
competition. If the immune system recognizes the escape mutant before its
fixation, the mutant might cease to be beneficial and disappear despite its
quick initial rise in frequency.  In support of this idea,
\citet{richman_rapid_2003, bunnik_autologous_2008} report antibody responses to
escape mutants. These respones are delayed by a few months, roughly matching the
average sweep time of an escape mutant. Alternatively, several different escape
mutations in the same epitope can arise almost simultaneously and start to
spread. Their fitness benefits are not additive, because each of them is
essentially sufficient to escape. As a consequence, several escape mutations rise to
high frequency, while the escape with the smallest cost in terms of replication,
packaging, etc. is most likely to
eventually fix. In simulations, this kind of epistatic interactions within
epitopes reduces the fixation probability. The emergence of
multiple sweeping nonsynonymous mutations in real HIV infections has been shown
previously~\citep{moore_limited_2009, bar_early_2012}.
See the supplementary material for examples of successful simulations in both scenarios.

@@

The observed hitchhiking highlights the importance of linkage due to infrequent
recombination for the evolution of HIV
\citep{neher_recombination_2010,batorsky_estimate_2011,
josefsson_majority_2011}. The recombination rate has been estimated to be on the
order of $\rho = 10^{-5}$ per base and day. It takes roughly $t_{sw} = s_a^{-1}
\log \nu_0$ generations for an adaptive mutation with growth rate $s_a$ to rise
from an initially low frequency $\nu_0\sim \mu$ to frequency one. This implies
that a region of length $l = (\rho t_{sw})^{-1} = s_a / \rho \log \nu_0$ remains
linked to the adaptive mutation. With $s_a=0.01$, $l\approx 100$ bases which is
consistent with strong linkage between the variable loops and the stems in
between. Furthermore, we do not expect hitchhiking to extend far beyond
the variable regions consistent with the lack of signal out side of C1-V5. In
case of much stronger selection -- such as observed during early CTL escape or
drug resistance evolution -- the linked  region is of course much larger.

The functional significance of the insulating RNA structure stems between the
hyper variable loops has been proposed
previously~\citep{watts_architecture_2009, sanjuan_interplay_2011}.
\citet{sanjuan_interplay_2011} have shown that insulating stems are relevant for
viral fitness {\it in vivo}. Our analysis is limited by the availability of
longitudinal data which requires a focus on the the variable regions of \env.
Conserved RNA structures most likely exist in different parts
of the HIV genome (several are known). In absence of repeated adaptive substitutions in the vicinity
that cause hitchhiking, the deleterious synonymous mutations remain at low
frequencies and can only be observed by deep sequencing methods.
